# Principal Component Analysis {#pca}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE, fig.width = 15, fig.height = 15)
#devtools::install_github("herveabdi/PTCA4CATA") # you only need this line of code if you haven't downloaded the package
#install.packages("gridExtra")
#install.packages("ggplotify")
#install.packages("grid")
library(devtools)
library(ExPosition)
library(corrplot)
library(ggplot2)
library(PTCA4CATA)
library(tidyverse)

# The following three functions allow you to arrange your plots in Rmarkdown
library(gridExtra)
library(ggplotify)
library(grid)
library(broom)
library(kableExtra)
```
## Method: Principal component analysis

*Principal component analysis* (PCA) is used to analyze one table of quantitative data. PCA mixes the input variables to give new variables, called principal components. The first principal component is the line of best fit. It is the line that maximizes the inertia (similar to variance) of the cloud of data points. Subsequent components are defined as orthogonal to previous components, and maximize the remaining inertia. 

PCA gives one map for the rows (called *factor scores*), and one map for the columns (called *loadings*). These 2 maps are related, because they both are described by the same components. However, these 2 maps project different kinds of information onto the components, and so they are *interpreted differently*. Factor scores are the coordinates of the row observations. They are interpreted by the distances between them, and their distance from the origin. Loadings describe the column variables. Loadings are interpreted by the angle between them, and their distance from the origin. 

The distance from the origin is important in both maps, because squared distance from the mean is inertia (variance, information; see sum of squares as in ANOVA/regression). Because of the Pythagorean Theorem, the total information contributed by a data point (its squared distance to the origin) is also equal to the sum of its squared factor scores. 

## Data set: Audio features

This is a dataset which describes audio features of songs in Spotify playlists. Specifically, the music.track dataset measures 165 songs on 16 variables, of which 11 are quantitative. Some of the audio features described are acousticness, danceability, and energy. 

```{r data_set, echo=FALSE}
df1 <- load("audio-feature.RData")
hwdata <- music.track
drop <- c("id", "name", "uri", "artist", "mode", "genre", "key", "time_signature")
df = hwdata[,!(names(hwdata) %in% drop)]

inputforscroll <- kable(head(df), col.names = c("Acousticness", "Danceability", "Duration in ms", "Energy", "Instrumentalness",
                                                   "Liveness", "Loudness", "Speechiness", "Tempo", "Valence"), align = 'c')
scroll_box(inputforscroll, height = "250px")

#kable(head(df))
```

### The correlation plot

In the correlation plot, we see some interesting, strong correlations. For instance, energy and acousticness are highly negatively correlated, just as loudness and acousticness. As expected, energy and loudness are strongly positively correlated, as are instrumentalness and acousticness. 

```{r correlation plot, results = T, echo = FALSE}
cor.res <- cor(df)
corrplot(cor.res, tl.cex = 1, tl.col = "black")
cor.plot <- recordPlot() # you need this line to be able to save the figure to PPT later
```

```{r analyze1, echo = TRUE}
res_pca <- epPCA(df, center = TRUE, scale = "SS1", DESIGN = hwdata$genre, graphs = FALSE)
```

### Analysis - Inference PCA

```{r InPosition, results = FALSE, echo = FALSE}
library(InPosition)
```
```{r, results = FALSE, echo = FALSE}
set.seed(42)
```

```{r analyze, results = FALSE, echo = T}
res_pcaInf <- epPCA.inference.battery(df, center = TRUE, scale = "SS1",
                                      DESIGN = music.track$genre, graphs = FALSE,
                                      test.iters = 999)
```

```{r showRes, results = FALSE, echo = FALSE}
res_pcaInf
```

```{r checkRes1, results = FALSE, echo = FALSE}
res_pcaInf$Fixed.Data
```

```{r, results = FALSE, echo = FALSE}
res_pcaInf$Inference.Data
```

### Permutation test plots 

The plot for Eigenvalue 1 shows that the observed value is very far from the 5% line (red dotted), hence implying that there is a possibility of the null being true less than 5% of the time.

```{r, results = FALSE, echo = FALSE}
zeDim = 1
pH2 <- prettyHist(
  distribution = res_pcaInf$Inference.Data$components$eigs.perm[,zeDim], 
           observed = res_pcaInf$Fixed.Data$ExPosition.Data$eigs[zeDim], 
           xlim = c(1, 6), # needs to be set by hand
           breaks = 20,
           border = "white", 
           main = paste0("Permutation Test for Eigenvalue ",zeDim),
           xlab = paste0("Eigenvalue ",zeDim), 
           ylab = "", 
           counts = FALSE, 
           cutoffs = c(0.975))
eigs1 <- recordPlot()
```

The permutation test for eigenvalue 2 does not show statistical significance


```{r, results = FALSE, echo = FALSE}
zeDim = 2
pH2 <- prettyHist(
  distribution = res_pcaInf$Inference.Data$components$eigs.perm[,zeDim], 
           observed = res_pcaInf$Fixed.Data$ExPosition.Data$eigs[zeDim], 
           xlim = c(1, 1.5), # needs to be set by hand
           breaks = 20,
           border = "white", 
           main = paste0("Permutation Test for Eigenvalue ",zeDim),
           xlab = paste0("Eigenvalue ",zeDim), 
           ylab = "", 
           counts = FALSE, 
           cutoffs = c(0.975))
eigs1 <- recordPlot()
```


### Scree Plot

The scree plot indicates that one component represents most of the variance in the dataset. The second and third components are quite close to each other, which warrants further investigation (but not at the cost of the first component).

```{r scree plot, results = FALSE, echo = FALSE}
factoextra::fviz_screeplot(res_pca, addlabels =TRUE, main = "PCA: Explained Variance per Dimension")
my.scree <- recordPlot() # you need this line to be able to save them in the end

```

### Row factor scores - inference PCA

The row factor scores plot shows the distribution of data, in this case, colored by different genres of music (sleep, dinner, party, and workout). 

Legend:

Pink = Sleep
Blue = Dinner
Green = Party
Cyan = Workout

The scores are distributed nicely and the components make sense (particularly component 1).

```{r factor_scores, results = FALSE, echo = FALSE}
my.fi.plot <- createFactorMap(res_pcaInf$Fixed.Data$ExPosition.Data$fi, # data
                            title = "Spotify Songs Row Factor Scores", # title of the plot
                            axis1 = 1, axis2 = 2, # which component for x and y axes
                            pch = 19, # the shape of the dots (google `pch`)
                            cex = 3, # the size of the dots
                            text.cex = 0, # the size of the text
                            alpha.points = 0.35,
                            display.labels = FALSE,
                            col.points = res_pcaInf$Fixed.Data$Plotting.Data$fi.col, # color of the dots
                            col.labels = res_pcaInf$Fixed.Data$Plotting.Data$fi.col, # color for labels of dots
                            )

fi.labels <- createxyLabels.gen(1,2,
                             lambda = res_pcaInf$Fixed.Data$ExPosition.Data$eigs,
                             tau = round(res_pcaInf$Fixed.Data$ExPosition.Data$t),
                             axisName = "Component "
                             )
fi.plot <- my.fi.plot$zeMap + fi.labels # you need this line to be able to save them in the end
fi.plot
```

```{r, results = FALSE, echo = FALSE}
# get index for the first row of each group
grp.ind <- order(music.track$genre)[!duplicated(sort(music.track$genre))]
grp.col <- res_pcaInf$Fixed.Data$Plotting.Data$fi.col[grp.ind] # get the color
grp.name <- music.track$genre[grp.ind] # get the corresponding groups
names(grp.col) <- grp.name
```

```{r fi_mean, results = FALSE, echo = FALSE}
group.mean <- aggregate(res_pcaInf$Fixed.Data$ExPosition.Data$fi,
                     by = list(music.track$genre), # must be a list
                     mean)
group.mean

# need to format the results from `aggregate` correctly
rownames(group.mean) <- group.mean[,1] # Use the first column as row names
fi.mean <- group.mean[,-1] # Exclude the first column
fi.mean
```

### Plotting the means

Adding means of the genre to the row factor scores plot helps in getting a clearer idea about where the centre of each group lies. Further, the similarity/difference between the genres can be understood by constructing confidence intervals around the means.  

```{r fimeanplot, results = FALSE, echo = FALSE}
fi.mean.plot <- createFactorMap(fi.mean,
                                alpha.points = 1,
                                col.points = grp.col[rownames(fi.mean)],
                                col.labels = grp.col[rownames(fi.mean)],
                                pch = 15,
                                cex = 4,
                                text.cex = 5)
fi.WithMean <- my.fi.plot$zeMap_background + my.fi.plot$zeMap_dots + 
  fi.mean.plot$zeMap_dots + fi.mean.plot$zeMap_text + fi.labels
fi.WithMean
```

### Tolerance interval

We can plot the tolerance interval for each genre.

```{r, results = FALSE, echo = FALSE}
TIplot <- MakeToleranceIntervals(res_pcaInf$Fixed.Data$ExPosition.Data$fi,
                            design = as.factor(music.track$genre),
                            col = grp.col[rownames(fi.mean)],
                            line.size = .50, 
                            line.type = 3,
                            alpha.ellipse = .2,
                            alpha.line    = .4,
                            p.level       = .95)
# If you get some errors with this function, check the names.of.factors argument in the help.

fi.WithMeanTI <- my.fi.plot$zeMap_background + my.fi.plot$zeMap_dots + 
  fi.mean.plot$zeMap_dots + fi.mean.plot$zeMap_text + TIplot + fi.labels

fi.WithMeanTI
```


### Bootstrap interval

We can also add the bootstrap interval for the group means to see if these group means are significantly different.

The ellipses around the group means indicates the bootstrap intervals. The smaller their radii, the more confidence we can assume of our group mean estimate. 

```{r, results='hide', echo = FALSE}
# Depend on the size of your data, this might take a while
fi.boot <- Boot4Mean(res_pcaInf$Fixed.Data$ExPosition.Data$fi,
                     design = music.track$genre,
                     niter = 1000)
# Check what you have
fi.boot

# What is the cube? Check the first 4 tables. You don't need to include this in 
# your output, because it's a lot of junk text.
fi.boot$BootCube[,,1:4]
```

```{r, results = FALSE, echo = FALSE}
# Check other parameters you can change for this function
bootCI4mean <- MakeCIEllipses(fi.boot$BootCube[,c(1:2),], # get the first two components
                              col = grp.col[rownames(fi.mean)])

fi.WithMeanCI <- my.fi.plot$zeMap_background + bootCI4mean + 
  my.fi.plot$zeMap_dots + fi.mean.plot$zeMap_dots + 
  fi.mean.plot$zeMap_text + fi.labels

fi.WithMeanCI

```

### Loadings

The circle of loadings shows the nature of correlations between the columns (variables).The interpretation is similar to that of plain PCA, with the angular distances indicative of the strnegth of correlation between two variables and the closer they are to the circumference, the higher their contribution to the dataset. 

```{r, results = FALSE, echo = FALSE}
cor.loading <- cor(df, res_pcaInf$Fixed.Data$ExPosition.Data$fi)
rownames(cor.loading) <- rownames(cor.loading)

cJ <- res_pca$ExPosition.Data$cj
color4Var <- prettyGraphs::prettyGraphsColorSelection(
  nrow(cJ))


loading.plot <- createFactorMap(cor.loading,
                                constraints = list(minx = -1, miny = -1,
                                                   maxx = 1, maxy = 1),
                                col.points = gplots::col2hex(color4Var),
                                col.labels = gplots::col2hex(color4Var))
LoadingMapWithCircles <- loading.plot$zeMap + 
  addArrows(cor.loading, color = gplots::col2hex(color4Var)) + 
  addCircleOfCor() + xlab("Component 1") + ylab("Component 2")

LoadingMapWithCircles
```

### Column factor scores

This plot shows the distribution of the variables in the PCA space. Once again looking at component 1, it is clear that party and workout songs are closely related to variables like danceability, valence, loudness, energy etc. On the other hand, dinner and sleep songs are correlated with instrumentalness and acousticness. 

```{r, results = FALSE, echo = FALSE}

## ----colors------------------------------------------
cJ <- res_pca$ExPosition.Data$cj
color4Var <- prettyGraphs::prettyGraphsColorSelection(
  nrow(cJ))

my.fj.plot <- createFactorMap(res_pcaInf$Fixed.Data$ExPosition.Data$fj, # data
                            title = "Spotify Songs Column Factor Scores", # title of the plot
                            axis1 = 1, axis2 = 2, # which component for x and y axes
                            pch = 19, # the shape of the dots (google `pch`)
                            cex = 3, # the size of the dots
                            text.cex = 3, # the size of the text
                            col.points = gplots::col2hex(color4Var), # color of the dots
                            col.labels = gplots::col2hex(color4Var), # color for labels of dots
                            )

fj.plot <- my.fj.plot$zeMap + fi.labels # you need this line to be able to save them in the end
fj.plot
```

### Contribution barplots and Bootstrap Ratio of columns

```{r, results = FALSE, echo = FALSE}
signed.ctrJ <- res_pcaInf$Fixed.Data$ExPosition.Data$cj * sign(res_pcaInf$Fixed.Data$ExPosition.Data$fj)

# plot contributions for component 1
ctrJ.1 <- PrettyBarPlot2(signed.ctrJ[,1],
                         threshold = 1 / NROW(signed.ctrJ),
                         font.size = 3,
                         ylab = 'Contributions',
                         ylim = c(1.2*min(signed.ctrJ[,1]), 1.2*max(signed.ctrJ[,1])),
                         horizontal = FALSE,
                         color4bar = gplots::col2hex(color4Var)
) + ggtitle("Contribution barplots", subtitle = 'Component 1: Variable Contributions (Signed)')

# plot contributions for component 2
ctrJ.2 <- PrettyBarPlot2(signed.ctrJ[,2],
                         threshold = 1 / NROW(signed.ctrJ),
                         font.size = 3,
                         color4bar = gplots::col2hex(color4Var), # we need hex code
                         ylab = 'Contributions',
                         ylim = c(1.2*min(signed.ctrJ[,2]), 1.2*max(signed.ctrJ[,2])),
                         horizontal = FALSE
) + ggtitle("",subtitle = 'Component 2: Variable Contributions (Signed)')


BR <- res_pcaInf$Inference.Data$fj.boots$tests$boot.ratios
laDim = 1

# Plot the bootstrap ratios for Dimension 1
ba001.BR1 <- PrettyBarPlot2(BR[,laDim],
                        threshold = 2,
                        font.size = 3,
                   color4bar = gplots::col2hex(color4Var), # we need hex code
                  ylab = 'Bootstrap ratios',
                  horizontal = FALSE
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle("Bootstrap ratios", subtitle = paste0('Component ', laDim))

# Plot the bootstrap ratios for Dimension 2
laDim = 2
ba002.BR2 <- PrettyBarPlot2(BR[,laDim],
                        threshold = 2,
                        font.size = 3,
                   color4bar = gplots::col2hex(color4Var), # we need hex code
                  ylab = 'Bootstrap ratios',
                  horizontal = FALSE,
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle("",subtitle = paste0('Component ', laDim))

```

```{r grid_ctrJ1, fig.width = 10, fig.height = 8, results = FALSE, echo = FALSE}
  grid.arrange(
    as.grob(ctrJ.1),
    as.grob(ctrJ.2),
    as.grob(ba001.BR1),
    as.grob(ba002.BR2),
    ncol = 2,nrow = 2,
    top = textGrob("Barplots for variables", gp = gpar(fontsize = 18, font = 3))
  )
BothCtrJ <- recordPlot() # you need this line to be able to save them in the end

print(BothCtrJ)
```

```{r, echo = FALSE}
# Here we can save all figures to a PowerPoint
# savedList <- saveGraph2pptx(file2Save.pptx = 'AllFigures_inf', 
#                             title = 'All Figures for inference', 
#                             addGraphNames = TRUE)

```

## Summary

When we interpret the factor scores and loadings together, the PCA revealed:

* Component 1: Sleep and dinner songs are different from party and workout songs. They differ on the characteristics of music such as danceability, loudness, valence, instrumentalness etc. 

Certain variables contribute more to the variance in the dataset than the others (acousticness, instrumentalness, valence, energy, danceability, loudness). 

Sleep and dinner songs are closely associated with acousticness and instrumentalness, while party and workout songs are explained more by valence, energy, danceability, and loudness. 