# Barycentric Discriminant Analysis {#bada}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE, fig.width = 15, fig.height = 15)
# Clean start
rm(list = ls())
graphics.off()
gc()
#_________________________________________________
# parameters for pptx ----
leTitre   <- 'BADA: Audio features'
leDir     <- paste0(getwd(),'/')    # Where am I
filename  <- 'BADA4AudioFeatures' # my title
path2save <-  paste0(leDir, filename)
#_________________________________________________
#_________________________________________________
# Preamble ----
# libraries ----
library(tidyverse)
library(ExPosition)
devtools::install_github("LukeMoraglia/ExPosition1/TExPosition")
devtools::install_github("LukeMoraglia/ExPosition1/TInPosition")
library(TExPosition)
library(TInPosition)
library(PTCA4CATA)
# devtools::install_github('HerveAbdi/data4PCCAR')
library(data4PCCAR)
library(kableExtra)
library(dplyr)
library(corrplot)
library(broom)
```

## Method: Barycentric Discriminant Analysis (BADA)

Discriminant analysis is focused on grouping observations into categories. Barycentric Discriminant Analysis is employed to analyze multiple variables describing a set of observations in which each observation belongs to one and only one category. The categories are predefined. 

BADA creates new combinations of the variables that best separate the groups in the data table. Using these new variables, a fresh observation measured using the same variables can placed in a category. The accuracy of placing the new observation into a category correctly is also calculated using cross validation techniques. 

Source - https://bit.ly/3k37ceO 

## Data set: Audio features
This is a dataset which describes audio features of songs in Spotify playlists. Specifically, the music.track dataset measures 165 songs on 16 variables, of which 11 are quantitative. Some of the audio features described are acousticness, danceability, and energy. 


```{r datasetBADA, echo=FALSE}
load("audio-feature.RData")
rawData <- music.track


                      
drop <- c("id")
rawData = rawData[,!(names(rawData) %in% drop)]
Xmat <-  rawData[, c(4, 5, 6, 7, 8, 10, 11, 13, 14, 16)]
Ymat <-  rawData[, c(9, 12, 15)]
XYmat <- rawData[, 4:16]

inputforscroll <- kable(XYmat)

scroll_box(inputforscroll, height = "250px")
```

### Analysis


```{r BADA, echo=TRUE}
# Run BADA  ----
resBADA <- tepBADA(XYmat, DESIGN = rawData$genre,
                   graphs = FALSE)
XYmat <- na.omit(XYmat)
```

```{r infBADA, echo=TRUE, results = 'hide'}
nIter = 1000
resBADA.inf <- tepBADA.inference.battery(XYmat, 
                  DESIGN = rawData$genre,
                  test.iters = nIter,
                  graphs = FALSE)
```

### Heatmap

The heatmap is simply a visual representation of the relationship between the variables and the genres of songs. As we can see, the most distinct relationship is between sleep songs and audio features of acousticness and instrumentalness (positive). On the other hand, sleep songs and energy have a negative relationship.  


```{r, echo=FALSE}
Xmat_scaled <- scale(Xmat, center = TRUE, scale = TRUE)

color4Var <- prettyGraphs::prettyGraphsColorSelection(
  ncol(XYmat))

CorrMeans <- PTCA4CATA::getMeans(Xmat_scaled, rawData$genre, FUN = mean)
corrplot(as.matrix(CorrMeans), is.corr = FALSE, addgrid.col = "black", tl.col = color4Var, tl.cex = 1, method = "color", col = colorRampPalette(c("violetred4", "white", "navyblue"))(200))

a0.heatmap <- recordPlot()

```


### The Scree plot

The scree plot shows us how many dimensions contribute to the variance in the data. In this plot, Dim 1 contributes more than 80% or the variance. Hence, it would be a good place to start. 

```{r, echo=FALSE}
# The ScreePlot. Fixed Effects. ----
# Get the ScreePlot
# scree for ev ----
PlotScree(ev = resBADA$TExPosition.Data$eigs,
   title = 'BADA Audio features: Inertia Scree Plot',
   plotKaiser = TRUE, 
   color4Kaiser = ggplot2::alpha('darkorchid4', .5),
   lwd4Kaiser  = 2)
# Save the plot for pptx
a0002.Scree.sv <- recordPlot()
```

```{r, echo=FALSE}
#Colors for the graphs
genreColors <- rawData$genre
# get colors for graphs
genreColors <- recode(genreColors, 
                     dinner = "#305ABF", 
                     party = '#84BF30', 
                     workout = '#BF30AD',
                     sleep = '#30BFA7')
```

### Map of row factors

This plot shows us the distribution of all elements in the data table. They are colored as per the genre, and we also see the group means marked. So pretty! 

```{r, echo=FALSE}
#  Observations and means ----
# Observations ----
#_________________________________________________
# I-set map ----
# a graph of the observations
Imap <- PTCA4CATA::createFactorMap(
  resBADA$TExPosition.Data$fii,
  col.points = genreColors,
  col.labels = genreColors,
  alpha.points = .35,
  text.cex = 0,
  display.labels = F,
  cex = 2
)

#_________________________________________________
# make labels ----
label4Map <- createxyLabels.gen(1,2,
          lambda = resBADA$TExPosition.Data$eigs,
          tau = resBADA$TExPosition.Data$t)
#_________________________________________________
# Fix Fi ----
# Note that current version of tepBADA 
# has a scaling problem
# Will soon be fixed
# wineMeans <- resBADA$TExPosition.Data$fi 
# code when BQDQ is fixed
audioMeans <- PTCA4CATA::getMeans(
               resBADA$TExPosition.Data$fii, 
               rawData$genre)
# a vector of color for the means
# Explicit recoding to make sure 
# that the names match
# => Needs to be improved!
col4Means <- recode(rownames(audioMeans), 
                      dinner = "#305ABF", 
                     party = '#84BF30', 
                     workout = '#BF30AD',
                     sleep = '#30BFA7')
names(col4Means) <- rownames(audioMeans)
# the map
MapGroup <- PTCA4CATA::createFactorMap(audioMeans,
             # use the constraint from the main map
             constraints = Imap$constraints,
             col.points = col4Means,
             cex = 4,  # size of the dot (bigger)
             col.labels = col4Means,
             text.cex = 4,
             display.labels = T,
             pch = 17,
             alpha.points = 1)
# The map with observations and group means
a003.bada <- Imap$zeMap + 
               label4Map +
               MapGroup$zeMap_dots + 
               MapGroup$zeMap_text
print(a003.bada)

```

### Confidence intervals generated by bootstrapping

The 95% confidence intervals around the group means are generated. They show the range of values the mean can assume 95% of the times the experiment is conducted, and the smaller the ellipsis, the higher confidence in the mean.

```{r, echo=FALSE}
# Confidence intervals
# Bootstrapped CI ----
#_________________________________________________
# Create Confidence Interval Plots
fi.boot <- resBADA.inf$Inference.Data$boot.data$fi.boot.data$boots
# We want to use the rownames of fi.boot 
# as reference to get the correct
# color. 
# However, the original rownames include "." 
# and don't match with 
# the original row names. 
# So, the `sub` function was used to get rid of 
# the "." by replacing all "." 
# in the rownames of fi.boot as an empty 
# string.
rownames(fi.boot) <- sub("[[:punct:]]","",
                          rownames(fi.boot))
# use function MakeCIEllipses 
# from package PTCA4CATA
GraphElli <- PTCA4CATA::MakeCIEllipses(
    resBADA.inf$Inference.Data$boot.data$fi.boot.data$boots,
    col = col4Means[rownames(fi.boot)], 
    # use rownames as reference to pick the color
    p.level = .95
)
#_________________________________________________
# create the I-map with Observations, 
# means and confidence intervals
#
a004.bada.withCI <-  Imap$zeMap_background  + 
                        Imap$zeMap_dots     + 
                        MapGroup$zeMap_dots + 
                        MapGroup$zeMap_text +
                      GraphElli + label4Map +
  ggtitle('BADA: Group Centers with CI and Observations')
#_________________________________________________
# plot it!
dev.new()
print(a004.bada.withCI)
```

### Row factors with hull

Hulls are drawn around each group. 

```{r, echo=FALSE}
# with Hull ----
Fii <- resBADA$TExPosition.Data$fii
colnames(Fii) <- paste0('Dimension ', 1:ncol(Fii))
# getting the color correct: an ugly trick
col4Hull <- col4Means[match(names(col4Means), 
                     levels(rawData$genre) )]
GraphHull <- PTCA4CATA::MakeToleranceIntervals(
                      Fii,
                      design = rawData$genre,
                      col = col4Hull,
                # the next line is required 
                # for some strange unknown reasons
          names.of.factors =  c("Dim1","Dim2"),
          p.level = 1.00)
#
a006.bada.withHull <-  Imap$zeMap_background  + 
                          Imap$zeMap_dots     + 
                          MapGroup$zeMap_dots + 
                          MapGroup$zeMap_text +
                          GraphHull           + 
                          label4Map           +
                          GraphElli +
    ggtitle('BADA: Group Centers with Hulls and Observations')
a006.bada.withHull

```

### Column factor scores

This plot shows us the correlations between the different variables - the angle between 2 arrows is an indication of the strength of relationship. The direction of relationship can be determined by looking at which quadrants the arrows lie in. 

```{r, echo=FALSE, results='hide'}
# J-set ----
# gt colors

cJ <- resBADA$TExPosition.Data$cj
color4Var <- prettyGraphs::prettyGraphsColorSelection(
  nrow(cJ))
col4X <- prettyGraphsColorSelection(
                   n.colors = ncol(Xmat),
                   starting.color = 42)

col4Y <- prettyGraphsColorSelection(
                    n.colors = ncol(Ymat),
                    starting.color = 13)
col4Var = c(col4X,col4Y)
col4Var <- as.matrix(col4Var)
#_________________________________________________
Fj <- resBADA$TExPosition.Data$fj
baseMap.j <- PTCA4CATA::createFactorMap(
                        Fj,
                        col.points   = col4Var,
                        alpha.points =  .3,
                        col.labels   = col4Var
                        )
print(baseMap.j)
#_________________________________________________
# arrows
zeArrows <- addArrows(Fj, color = col4Var)
# A graph for the J-set
# A graph for the J-set
b001.aggMap.j <- baseMap.j$zeMap_background + # background layer
                      baseMap.j$zeMap_dots + 
                      baseMap.j$zeMap_text +  # dots & labels
                      label4Map 
b002.aggMap.j <- b001.aggMap.j + zeArrows
# We print this Map with the following code
#dev.new()
print(b002.aggMap.j)

```


### Contribution bar plots for dimension 1

The contribution bar plots show how much each variable is contributing to a certain dimension. Bootstrap ratios that follow are derived from the inference battery. 

```{r, echo=FALSE}
#  Contributions ####
#_________________________________________________
#_________________________________________________
# Ctr J-set 
###### 1 ====
# 
ctrj <- resBADA$TExPosition.Data$cj
signed.ctrj <- ctrj * sign(Fj)
# BR1
c001.plotCtrj.1 <- PrettyBarPlot2(
           bootratio = round(100*signed.ctrj[,1]), 
           threshold = 100 / nrow(signed.ctrj), 
           ylim = NULL, 
           color4bar = gplots::col2hex(col4Var),
           color4ns = "gray75", 
           plotnames = TRUE, 
           horizontal = F,
    main = 'Important Contributions Variables. Dim 1.', 
           ylab = "Signed Contributions")

print(c001.plotCtrj.1)

```


### Contribution bar plots for dimension 2

```{r, echo=FALSE}
###### 2 ====
# 
c002.plotCtrj.2 <- PrettyBarPlot2(
  bootratio = round(100*signed.ctrj[,2]), 
  threshold = 100 / nrow(signed.ctrj), 
  ylim = NULL, 
  color4bar = gplots::col2hex(col4Var),
  color4ns = "gray75", 
  plotnames = TRUE, 
  horizontal = F,
  main = 'Important Contributions Variables. Dim 2.', 
  ylab = "Signed Contributions")

print(c002.plotCtrj.2)

```


### Bootstrap ratio - Dimension 1

```{r, echo=FALSE}
#  Bootstrap ratios ####
#_________________________________________________
#_________________________________________________
#BR. 1 ====
# 
BRj <- resBADA.inf$Inference.Data$boot.data$fj.boot.data$tests$boot.ratios
# BR1
d001.plotBRj.1 <- PrettyBarPlot2(
  bootratio = BRj[,1], 
  threshold = 2, 
  ylim = NULL, 
  color4bar = gplots::col2hex(col4Var),
  color4ns = "gray75", 
  plotnames = TRUE, 
  horizontal = F,
  main = 'Bootstrap Ratios Variables. Dim 1.', 
  ylab = "Bootstrap Ratios")

print(d001.plotBRj.1)

```


### Bootstrap ratio - Dimension 2

```{r, echo=FALSE}
###### 2 ====
# 
d003.plotBRj.2 <- PrettyBarPlot2(
  bootratio = BRj[,2], 
  threshold = 2, 
  ylim = NULL, 
  color4bar = gplots::col2hex(col4Var),
  color4ns = "gray75", 
  plotnames = TRUE, 
  horizontal = F,
  main = 'Bootstrap Ratios Variables. Dim 2.', 
  ylab = "Bootstrap Ratios")

print(d003.plotBRj.2)
```

### Confusion matrix - fixed effects

The fixed effects confusion matrix gives us an idea of how good our predictions were to create the categories in the existing data. The accuracy value quantifies the same. On the other hand, random effects confusion matrix outlines the quality of prediction/categorizing accuracy of _new_ observations in the dataset.  

*Jackknife* is a technique used to train models for prediction by separating the dataset of interest into training and testing data. Testing data will contain only one row/subject/participant/observation, which will be predicted using the attributes of the rest of the data. 


```{r, echo=FALSE}

confusionmatrix1 <- resBADA.inf$Inference.Data$loo.data$fixed.confuse

resBADA.inf$Inference.Data$loo.data$fixed.acc

kable(as.data.frame(confusionmatrix1))

```

### Confusion matrix - random effects

```{r, echo=FALSE}

confusionmatrix <- resBADA.inf$Inference.Data$loo.data$loo.confuse

resBADA.inf$Inference.Data$loo.data$loo.acc

kable(as.data.frame(confusionmatrix))

```


```{r, echo=FALSE}
# savedList <- PTCA4CATA::saveGraph2pptx(
#                   file2Save.pptx = path2save, 
#                   title = leTitre, 
#                   addGraphNames = TRUE)
```

## Summary:
Dimension 1
Rows: 		Dinner and Sleep	VS 	Party and Workout
Cols: 		Danceability, loudness, energy are highly positively correlated, acousticness and energy strongly negatively correlated

Dimension 2
Sleep and Party/Workout songs are distinct from each other.
Duration contributes to the dimension more than others. Seems to have an impact on dinner songs.

Interpretation: Dinner and sleep music that are more acoustical/instrumental are distinct from other genres
